Manuscript Figures

### 2021-03-25 src/manuscript_figures.Rmd
update survival analysis - use RSF to filter variables

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, include = FALSE, warning = FALSE, message = FALSE)
# todo: pass this as argument
settings <- yaml::read_yaml('/data/local/buyar/arcas/pancancer_multiomics_manuscript/settings.yaml')
source(settings$utility_script)
library(data.table)
library(ggplot2)
library(pbapply)
library(ggrepel)
library(ggpubr)
library(cowplot)
library(ggridges)
library(ggcorrplot)
library(ComplexHeatmap)
library(pheatmap)
theme_set(ggpubr::theme_pubclean())
geneSetScoresFolder <- '/data/local/buyar/arcas/subtyping_paper/data/geneset_scores/singscore'
FigureFolder <- '/data/bimsbstatic/public/akalin/buyar/manuscript_figures_arcas'

@@

# do unsupervised clustering and compute AMI 
experiment <- 'pancancer.cnv_gex_meth_mut'
df <- do.call(rbind, pbapply::pblapply(TOOLS, function(x) {
  message(x)
  M <- LFs[[x]][[experiment]]
  if(x == 'maui_factors') {
    M <- remove_redundant_variables(M, perc = 99)
  }
  # find cancer types per sample
  ct <- surv[match(rownames(M), bcr_patient_barcode)]$project
  data.frame('tool' = x, 
             'AMI' = compute_AMI(M, k = length(unique(ct)), target_labels = ct))
}))
df$tool <- toupper(gsub("_factors", "", df$tool))
p5 <- ggplot(df, aes(x = tool, y = AMI)) +
  geom_bar(aes(fill = tool), stat = 'identity') + 
  geom_text(aes(label = round(AMI, 2))) + 
  theme(legend.title = element_blank(), axis.title.x = element_blank())

@@

    # try topN from 5 to 25 (each tool may capture different amount of information 
    # by varying number of factors)
    stats <- do.call(rbind, lapply(seq(5, min(25, ncol(M)), 5), function(n) {
      message(n)
      M.sub <- t(subset_features_by_variance(t(M), topN = n))
      val <- c('internal') 
      if(ncol(M) > 2) {val <- c('internal', 'stability')}
      res <- clValid::clValid(obj = M.sub,
                            nClust = 3:8,
                            clMethods = 'kmeans',
                            validation = val,
                            maxitems = 10000,
                            nstart = 1000, iter.max = 30)
      stats <- data.table::as.data.table(res@measures)[V1 %in% c('Silhouette', 'APN')]
      colnames(stats) <- c('metric', 'k', 'clustering_method', 'score')
      stats$top_factors <- n
      return(stats)
    }))
    stats$cancer_type <- strsplit(experiment, "\\.")[[1]][1]
    stats$datatypes <- strsplit(experiment, "\\.")[[1]][2]
    return(stats)
  }))
  stats$method <- tool
  return(stats)
}))
parallel::stopCluster(cl)

@@

# combine both scores and plot for different values of k
# dt <- dcast(cluster.stats[datatypes == multi_omics_comb], 
#             cancer_type + method + k ~ metric, value.var = 'score')
# dt$kv <- paste0('kmeans(k=',dt$k,')')
# dt$method <- toupper(gsub("_factors", "", dt$method))
# p3 <- ggplot(dt,
#        aes(x = Silhouette, y = APN)) + 
#         geom_point(aes(color = method), size = 3) + 
#         facet_wrap(~ kv, ncol = 2) +
#         labs(title = 'Multi-omics clustering: Silhouette vs APN')

# survival prediction accuracy benchmarking
# find survival-predictive LFs by univariate cox models and use them in RSF
cl <- parallel::makeCluster(20)
parallel::clusterExport(cl, varlist = c('LFs', 'experiments', 'settings', 'surv'))
surv.stats <- do.call(rbind, sapply(simplify = F, names(LFs), function(tool) {
  message("survival benchmark for ",tool)
  do.call(rbind, pbapply::pbsapply(cl = cl, simplify = F, experiments, function(experiment) {
    message(experiment)
    source(settings$utility_script)
    require(data.table)
    require(survminer)
    M <- LFs[[tool]][[experiment]]
    if(tool == 'maui_factors') {
      M <- remove_redundant_variables(M, 99)
    }
    dt <- data.table('cindex.pfi' = NA, 'cindex.os' = NA, 'experiment' = experiment, 'method' = tool)
    # compute pfi stats
    vars <- get_prognostic_covariates(surv, clin = NULL, explored_covariates = M,
                                      event_col = 'PFI', duration_col = 'PFI.time')
    if(!is.null(vars)) {
      v <- head(vars[order(pval)][pval < 0.05][['variable']], 10)
      fit <- fit_RSF(surv, M[,v], 1000, 'PFI.time', 'PFI')
      dt$cindex.pfi <- round(1 - mean(fit$err.rate, na.rm = T), 2)
    }
    vars <- get_prognostic_covariates(surv, clin = NULL, explored_covariates = M,
                                      event_col = 'OS', duration_col = 'OS.time')
    if(!is.null(vars)) {
      v <- head(vars[order(pval)][pval < 0.05][['variable']], 10)
      fit <- fit_RSF(surv, M[,v], 1000, 'OS.time', 'OS')
      dt$cindex.os <- round(1 - mean(fit$err.rate, na.rm = T), 2)
    }
    return(dt)
  }))
}))
parallel::stopCluster(cl)

dt <- melt(surv.stats, measure.vars = c('cindex.pfi', 'cindex.os'))
dt[is.na(value)]$value <- min(dt$value, na.rm = T) # 
dt$method <- toupper(gsub("_factors", "", dt$method))
dt$project <- unlist(sapply(strsplit(dt$experiment, "\\."), function(x) x[1]))
dt$endpoint <- ifelse(dt$variable == 'cindex.pfi', 'Progression Free Survival', 
                      'Overall Survival')
# PCA/MOFA win, compare against PCA
comps <- lapply(setdiff(unique(dt$method), 'MAUI'), function(x) c('MAUI', x))
p4 <- ggboxplot(dt, x = 'method', y = 'value', color = 'method',
                add = 'jitter', facet.by = 'endpoint') +
  stat_compare_means(method = 'wilcox.test', paired = TRUE,
                     comparisons = comps) +
  labs(y = "Harrel's C-index", x = '') +
  theme(legend.position = 'none')

@@

experiments <- grep('^TCGA-.*.cnv_gex_meth_mut', names(LFs$maui_factors), value = T)
cl <- parallel::makeCluster(20)
parallel::clusterExport(cl, varlist = c('LFs', 'clin_factors', 
                                        'surv', 'experiments', 'settings'))
surv.stats <- pbapply::pbsapply(cl = cl, simplify = F, experiments, function(experiment) {
    require(data.table)
    source(settings$utility_script)
    message(experiment)
    M <- LFs$maui_factors[[experiment]]
    # get clinical factors for the patients in matrix
    clin <- process_clinical_covariates(rownames(M))
    clin <- subset(clin, select = intersect(colnames(clin), clin_factors))
    res <- get_prognostic_covariates(surv, clin, M, 'PFI', 'PFI.time')
    if(!is.null(res)) {
      cancer_type <- strsplit(experiment, "\\.")[[1]][1]
      res$project <- cancer_type
      v <- res[order(pval)][1]$variable
      dat <- merge(clin, M[,v,drop = F], by = 'row.names')
      rownames(dat) <- dat$Row.names
      dat$Row.names <- NULL
      dat <- get_surv_df(surv, dat) # add PFI values
      # compute hazard ratios 
      cox1 <- coxph(Surv(PFI.time,PFI) ~ ., 
              data=dat, x = TRUE)
      hr <- data.frame(summary(cox1)[['conf.int']]) # get hazard ratios with conf intervals
      hr <- hr[grep(v, rownames(hr)), c(1,3,4)]
      colnames(hr) <- c('hr', 'hr_low', 'hr_high')
      rownames(hr) <- v
      p1 <- survminer::ggforest(cox1, data = dat)

      # split variable into high/low to compute survival adjusted curves
      cp <- find_survival_cutpoint(dat[,v,drop=F], surv)
      dat[,v] <- ifelse(dat[,v] > cp, paste(v, 'high'), paste(v, 'low'))
      cox2 <- coxph(Surv(PFI.time,PFI) ~ ., 
                    data=dat, x = TRUE)
      p2 <- survminer::ggadjustedcurves(fit = cox2, data = dat, variable = v, method = 'average')

      # compute Cindex with all survival-predictive LFs and without LFs 
      vars <- res[pval < 0.05]$variable
      rsf_stats <- fit_RSF_clin(get_surv_df(surv, clin), M[,vars], ntree = 1000)

      return(list('stats' = res, 'p1' = p1, 'p2' = p2, 'hr' = hr, 
                  'Cindex_base' = rsf_stats$C_wo, 'Cindex_with_LFs' = rsf_stats$C_with))
    }
    return(NULL)
  })
parallel::stopCluster(cl)

@@

experiments <- grep('^TCGA-.*.cnv_gex_meth_mut', names(LFs$maui_factors), value = T)
cl <- parallel::makeCluster(20)
parallel::clusterExport(cl, varlist = c('LFs', 'clin_factors', 'surv', 
                                        'experiments', 'settings'))
prog.clust <- pbapply::pbsapply(cl = cl, simplify = F, experiments, 
                                function(experiment){
    require(data.table)
    require(survminer)
    source(settings$utility_script)
    message(experiment)
    M <- LFs$maui_factors[[experiment]]
    # get clinical factors for the patients in matrix
    clin <- process_clinical_covariates(rownames(M))
    clin <- subset(clin, select = intersect(colnames(clin), clin_factors))
    res <- get_prognostic_covariates(surv, clin, M, 'PFI', 'PFI.time')
    # now cluster samples based on only survival predictive factors
    prog_lfs <- unique(res$variable)
    if(length(prog_lfs) < 2) {
      return(NULL)
    }
    clust <-  get_kmeans_subtypes(M[,prog_lfs], k_vals = 3:8, min_cluster_size = 0)
    # find value of k which minimizes log-rank p-value
    dt <- do.call(rbind, lapply(colnames(clust), function(x) {
      message(x)
      pval <- plot_survival(surv, clust[,x,drop=F], return_pval = T)
      data.table('k' = x, 'pval' = pval)
    }))

@@

Fig1 <- cowplot::plot_grid(plotlist = bipartite_plots, nrow = 3, labels = 'AUTO')
ggsave('test1.pdf', plot = Fig1, units = 'in', width = 12, height = 14)

Fig <- cowplot::plot_grid(p1, p2, labels = 'AUTO', nrow = 2)
#ggsave('test2.pdf', plot = Fig2, units = 'in', width = 14, height = 12)

supFig_count <- supFig_count + 1
SuppFigures[[supFig_count]] <- Fig
fig.name <- paste0('SuppFigure', supFig_count, '.subtype_dependency.pdf')
names(SuppFigures)[supFig_count] <- fig.name
ggsave(file.path(FigureFolder, fig.name), 
              plot = Fig, units = 'in', width = 18, height = 12)
SuppLegends[[supFig_count]] <- "Demonstration of how cancer subtypes are context/objective dependent.
A) Heatmap of patient cluster memberships based on different objectives: optimum stratification based on overall survival or progression free survival, molecular homogeneity (TCGA pan-GI subtypes), or micro-satellite instability (MSI) status. Columns are ordered first by membership in the optimum Overall Survival cluster membership. B) Pairwise Adjusted Mutual Information (AMI) of cluster labels obtained via different patient stratification methods. The AMI value ranges from 0 to 1 and the higher the value, the more overlap between patient partitions."
```

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, include = FALSE, warning = FALSE, message = FALSE)
# todo: pass this as argument
settings <- yaml::read_yaml('/data/local/buyar/arcas/pancancer_multiomics_manuscript/results/hallmark_genes/settings.yaml')
source(settings$utility_script)
library(data.table)
library(ggplot2)
library(pbapply)
library(ggrepel)
library(ggpubr)
library(cowplot)
library(ggridges)
library(ggcorrplot)
library(ComplexHeatmap)
library(pheatmap)
theme_set(ggpubr::theme_pubclean())
geneSetScoresFolder <- '/data/local/buyar/arcas/subtyping_paper/data/geneset_scores/singscore'
FigureFolder <- '/data/bimsbstatic/public/akalin/buyar/manuscript_figures_arcas/test'

@@

# do unsupervised clustering and compute AMI 
experiment <- 'pancancer.cnv_gex_meth_mut'
df <- do.call(rbind, pbapply::pblapply(TOOLS, function(x) {
  message(x)
  M <- LFs[[x]][[experiment]]
  if(x == 'maui_factors') {
    #M <- remove_redundant_variables(M, perc = 99)
    M <- scale(M)
  }
  # find cancer types per sample
  ct <- surv[match(rownames(M), bcr_patient_barcode)]$project
  data.frame('tool' = x, 
             'AMI' = compute_AMI(M, k = length(unique(ct)), target_labels = ct))
}))
df$tool <- toupper(gsub("_factors", "", df$tool))
p5 <- ggplot(df, aes(x = tool, y = AMI)) +
  geom_bar(aes(fill = tool), stat = 'identity') + 
  geom_text(aes(label = round(AMI, 2))) + 
  theme(legend.title = element_blank(), axis.title.x = element_blank())

@@

    # try topN from 5 to 25 (each tool may capture different amount of information 
    # by varying number of factors)
    stats <- do.call(rbind, lapply(seq(5, min(20, ncol(M)), 5), function(n) {
      message(n)
      M.sub <- t(subset_features_by_variance(t(M), topN = n))
      val <- c('internal') 
      if(ncol(M.sub) > 2) {val <- c('internal', 'stability')}
      res <- clValid::clValid(obj = M.sub,
                            nClust = 3:8,
                            clMethods = 'kmeans',
                            validation = val,
                            maxitems = 10000,
                            nstart = 1000, iter.max = 30)
      stats <- data.table::as.data.table(res@measures)[V1 %in% c('Silhouette', 'APN')]
      colnames(stats) <- c('metric', 'k', 'clustering_method', 'score')
      stats$top_factors <- n
      return(stats)
    }))
    stats$cancer_type <- strsplit(experiment, "\\.")[[1]][1]
    stats$datatypes <- strsplit(experiment, "\\.")[[1]][2]
    return(stats)
  }))
  stats$method <- tool
  return(stats)
}))
parallel::stopCluster(cl)

@@

# survival prediction accuracy benchmarking
# find survival-predictive LFs by univariate cox models and use them in RSF
clin_factors <- c('age_at_initial_pathologic_diagnosis', 'gender', 'ajcc_pathologic_tumor_stage')
cl <- parallel::makeCluster(20)
parallel::clusterExport(cl, varlist = c('LFs', 'experiments', 'settings', 'surv', 'clin_factors'))
surv.stats <- do.call(rbind, sapply(simplify = F, names(LFs), function(tool) {
  message("survival benchmark for ",tool)
  do.call(rbind, pbapply::pbsapply(cl = cl, simplify = F, experiments, function(experiment) {
    message(experiment)
    source(settings$utility_script)
    require(data.table)
    require(survminer)
    M <- LFs[[tool]][[experiment]]
    clin <- process_clinical_covariates(rownames(M))
    clin <- subset(clin, select = intersect(colnames(clin), clin_factors))
    # compute pfi stats
    # fit a first model to filter variables 
    fit <- fit_RSF(surv, M, ntree = 500, 'PFI.time', 'PFI')
    # using a range of top variables, compute c index
    dt1 <- do.call(rbind, lapply(2:10, function(i) {
      vars <- names(head(sort(fit$importance, decreasing = T), i))
      rsf_stats <- fit_RSF_clin(get_surv_df(surv, clin, 'PFI', 'PFI.time'), 
                                M[,vars], ntree = 500)
      dt <- data.table('cindex.base' = rsf_stats$C_wo, 'cindex' = rsf_stats$C_with, 
                       'experiment' = experiment, 'method' = tool, 'topN' = i)
    }))
    dt1$endpoint <- 'PFI'
    # do the same for overall survival
    fit <- fit_RSF(surv, M, ntree = 500, 'OS.time', 'OS')
    dt2 <- do.call(rbind, lapply(2:10, function(i) {
      vars <- names(head(sort(fit$importance, decreasing = T), i))
      rsf_stats <- fit_RSF_clin(get_surv_df(surv, clin, 'OS', 'OS.time'), 
                                M[,vars], ntree = 500)
      dt <- data.table('cindex.base' = rsf_stats$C_wo, 'cindex' = rsf_stats$C_with, 
                       'experiment' = experiment, 'method' = tool, 'topN' = i)
    }))
    dt2$endpoint <- 'OS'
    return(rbind(dt1, dt2))
  }))
}))
parallel::stopCluster(cl)

dt <- surv.stats
dt$method <- toupper(gsub("_factors", "", dt$method))
dt$project <- unlist(sapply(strsplit(dt$experiment, "\\."), function(x) x[1]))
dt$endpoint <- ifelse(dt$endpoint == 'PFI', 'Progression Free Survival', 
                      'Overall Survival')
# get best performance per tool per project
dt <- dt[,.SD[which.max(cindex)], by = c('project', 'method', 'endpoint')]
# PCA/MOFA win, compare against PCA
comps <- lapply(setdiff(unique(dt$method), 'MAUI'), function(x) c('MAUI', x))
p4 <- ggboxplot(dt, x = 'method', y = 'cindex', color = 'method',
                add = 'jitter', facet.by = 'endpoint') +
  stat_compare_means(method = 'wilcox.test', paired = TRUE,
                     comparisons = comps) +
  labs(y = "Harrel's C-index", x = '') +
  theme(legend.position = 'none')

@@

# factors to use in the model
clin_factors <- c('age_at_initial_pathologic_diagnosis', 'gender', 'ajcc_pathologic_tumor_stage')
event <- 'PFI'
time <- 'PFI.time'

experiments <- grep('^TCGA-.*.cnv_gex_meth_mut', names(LFs$maui_factors), value = T)
cl <- parallel::makeCluster(20)
parallel::clusterExport(cl, varlist = c('LFs', 'clin_factors', 'event', 'time',
                                        'surv', 'experiments', 'settings'))
surv.stats <- pbapply::pbsapply(cl = cl, simplify = F, experiments, function(experiment) {
  require(data.table)
  require(survminer)
  require(survival)
  source(settings$utility_script)
  message(experiment)
  M <- LFs$maui_factors[[experiment]]
  # get clinical factors for the patients in matrix
  clin <- process_clinical_covariates(rownames(M))
  clin <- subset(clin, select = intersect(colnames(clin), clin_factors))
  # first filter variables by importance 
  fit <- fit_RSF(surv, M, 500, time, event)
  vars <- names(head(sort(fit$importance, decreasing = T), 5))

  # compute hazard ratios for the top factor
  v <- vars[1] 
  dat <- merge(clin, M[,v,drop = F], by = 'row.names')
  rownames(dat) <- dat$Row.names
  dat$Row.names <- NULL
  dat <- get_surv_df(surv, dat, event, time)
  # compute hazard ratios 
  cox1 <- coxph(Surv(time,status) ~ ., 
          data=dat, x = TRUE)
  hr <- data.frame(summary(cox1)[['conf.int']]) # get hazard ratios with conf intervals
  hr <- hr[grep(v, rownames(hr)), c(1,3,4)]
  colnames(hr) <- c('hr', 'hr_low', 'hr_high')
  rownames(hr) <- v
  p1 <- survminer::ggforest(cox1, data = dat)

  # split variable into high/low to compute survival adjusted curves
  cp <- find_survival_cutpoint(dat[,v,drop=F], surv, event_col = event, duration_col = time)
  dat[,v] <- ifelse(dat[,v] > cp, paste(v, 'high'), paste(v, 'low'))
  cox2 <- coxph(Surv(time,status) ~ ., 
                data=dat, x = TRUE)
  p2 <- survminer::ggadjustedcurves(fit = cox2, data = dat, variable = v, method = 'average')

  # compute Cindex with all survival-predictive LFs and without LFs 
  rsf_stats <- fit_RSF_clin(get_surv_df(surv, clin, event, time), 
                            M[,vars], ntree = 1000)

  return(list('p1' = p1, 'p2' = p2, 'hr' = hr, 
              'Cindex_base' = rsf_stats$C_wo, 'Cindex_with_LFs' = rsf_stats$C_with))

  return(NULL)
})
parallel::stopCluster(cl)

@@

experiments <- grep('^TCGA-.*.cnv_gex_meth_mut', names(LFs$maui_factors), value = T)
cl <- parallel::makeCluster(20)
parallel::clusterExport(cl, varlist = c('LFs', 'clin_factors', 'surv', 
                                        'experiments', 'settings'))
prog.clust <- pbapply::pbsapply(cl = cl, simplify = F, experiments, 
                                function(experiment){
    require(data.table)
    require(survminer)
    source(settings$utility_script)
    message(experiment)
    M <- LFs$maui_factors[[experiment]]
    # get clinical factors for the patients in matrix
    clin <- process_clinical_covariates(rownames(M))
    clin <- subset(clin, select = intersect(colnames(clin), clin_factors))

    fit <- fit_RSF(surv, M, duration_col = 'PFI.time', event_col = 'PFI')
    prog_lfs <- names(head(sort(fit$importance, decreasing = T), 5))
    # now cluster samples based on only survival predictive factors
    clust <-  get_kmeans_subtypes(M[,prog_lfs], k_vals = 3:8, min_cluster_size = 0)
    # find value of k which minimizes log-rank p-value
    dt <- do.call(rbind, lapply(colnames(clust), function(x) {
      message(x)
      pval <- plot_survival(surv, clust[,x,drop=F], return_pval = T)
      data.table('k' = x, 'pval' = pval)
    }))

@@

Fig <- cowplot::plot_grid(p1, p2, labels = 'AUTO', nrow = 2)
supFig_count <- supFig_count + 1
SuppFigures[[supFig_count]] <- Fig
fig.name <- paste0('SuppFigure', supFig_count, '.subtype_dependency.pdf')
names(SuppFigures)[supFig_count] <- fig.name
ggsave(file.path(FigureFolder, fig.name), 
              plot = Fig, units = 'in', width = 18, height = 12)
SuppLegends[[supFig_count]] <- "Demonstration of how cancer subtypes are context/objective dependent.
A) Heatmap of patient cluster memberships based on different objectives: optimum stratification based on overall survival or progression free survival, molecular homogeneity (TCGA pan-GI subtypes), or micro-satellite instability (MSI) status. Columns are ordered first by membership in the optimum Overall Survival cluster membership. B) Pairwise Adjusted Mutual Information (AMI) of cluster labels obtained via different patient stratification methods. The AMI value ranges from 0 to 1 and the higher the value, the more overlap between patient partitions."

Fig <- cowplot::plot_grid(plotlist = bipartite_plots, nrow = 3, labels = 'AUTO')
supFig_count <- supFig_count + 1
SuppFigures[[supFig_count]] <- Fig
fig.name <- paste0('SuppFigure', supFig_count, '.subtype_dependency.bipartite_plots.pdf')
names(SuppFigures)[supFig_count] <- fig.name
ggsave(file.path(FigureFolder, fig.name), 
              plot = Fig, units = 'in', width = 14, height = 14)
SuppLegends[[supFig_count]] <- "Demonstration of how cancer subtypes are context/objective dependent"
```
